Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-05-07 paper/Paper.Rnw
moved a subsection and a figure that goes with it to sup mat to save length and complexity

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the rare/common species growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$. We will refer to this term as 
the \emph{ATA effect on coexistence}. 
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

\subsection{Removing asymmetries of association}

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? Fig. \ref{fig:pedagog2} provides an illustration of the steps we describe here. 
We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal (Fig. \ref{fig:pedagog2}a,b).  
Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\E(e_i)=\E(c_{i \bs i})=0$ and $\var(e_i)=\var(c_{i \bs i})=1$, and we denote 
$\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$ (Fig. \ref{fig:pedagog2}c). We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$ (Fig. \ref{fig:pedagog2}d). 
The value of $\ddot{\rho}_i$ is then determined in one of 
four alternative ways: either by setting $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated.
For those familiar with copula statistics (see, e.g., \citealp{Genest_2007} and \citealp{Ghosh_2020_AER}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created similarly.

@@

%***DAN: This looks pretty good except one of the histograms on c does not look like the plotted
%pdfs of the marginals. I think Jasmin said in an email she was fixing this.
%***DAN: Jasmin, also, we use \varphi in the text, but \phi here. These need to be made consistent, since they are different to
%a math person. The easiest way will be to change it in the figure, so I hope that can straightforwardly be done.
\begin{figure}
\includegraphics[width=1.025\textwidth]{../results_figs/theory_fig_col.pdf}
\caption{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATA present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{||}$ and  $C_{i \bs i}^{||}$,
%Dan replaced \nat in the above two expressions by || because for some reason the \nat does not
%compile when it occurs within a caption. Not sure how to deal with that, but leaving it alone for now.
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
\label{fig:pedagog2}}
\end{figure}

>>>>

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the rare/common species growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of 
tail association are removed -- see SI section \ref{SI-sect:remove_asymmetries} for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$. We will refer to this term as 
the \emph{ATA effect on coexistence}. 
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

@@

### 2023-05-07 paper/SupMat.Rnw
cleaning up a bit is all
Dan finished going through the paper, about to try cutting 550 words
moved a subsection and a figure that goes with it to sup mat to save length and complexity

@@

%***DAN: I think I may need the below later, in the computation section, but not here
%For speed of computation, when samples $(b_1^{(i)},b_2^{(i)})$, $i=1,...,M$ were needed for distinct 
%sets of values of the parameters $\mu_1$, $\mu_2$, and $\sigma$, we used the same 
%$\left(\tilde b_1^{(i)},\tilde b_2^{(i)}\right)$, $i=1,...,M$, simply transforming these values 
%differently (using the different valus of $\mu_1$, $\mu_2$, $\sigma$) for each set of parameter values.

@@

%***DAN: We might need this text somewhere, but not sure where. Just saving it out of caution.
% For the lottery model, recall that $E_i=B_i$ and $C_{i \bs i} = B_j/\delta$. So 
% $\varphi^{-1} \circ F_{E_i}(E_i) = \frac{b_i-\mu_i}{\sigma}$ and 
% $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})=\frac{b_j - \mu_j}{\sigma}$.  
% Also, $\rho_i = \cov(\frac{b_i-\mu_i}{\sigma},\frac{b_j-\mu_j}{\sigma})$, which equals the 
% Pearson correlation of $b_i$ and $b_j$. For the lottery model, it seems natural to make the choice
% $\ddot{\rho}_i=\rho_i$, which straightforwardly implies that $(E_i^\nat,C_{i \bs i}^\nat)$
% can be written as $(\exp(\ddot{b}_i),\exp(\ddot{b}_j))$, where $(\ddot{b}_i,\ddot{b}_j)$
% is bivariate-normally distributed with mean $(\mu_i,\mu_j)$, $\var(\ddot{b}_i)=\var(\ddot{b}_j)=\sigma$,
% and $\cov(\ddot{b}_i,\ddot{b}_j)=\cov(b_i,b_j)$. So $(\ddot{b}_i,\ddot{b}_j)$ has the same
% marginal distributions and covariance matrix as $(b_i,b_j)$, but it is bivariate normal, whereas
% $(b_i,b_j)$ itself was not required to be. Given these definitions, it is straightforward to compute the 
% component $\olrn_{i \bs i}$ of main text equation (\ref{MT-eq:storage_effects_further1}) for the lottery model using the simulation 
% approach of \cite{Ellner_2016,Ellner_2019}. Note that $E_j$ and $C_{j \bs i}$ are linearly related for the lottery model,
% so $E_j^\nat=E_j$, $C_{j \bs i}^\nat = C_{j \bs i}$, and $\olrn_{j \bs i}=\olr_{j \bs i}$.

@@

%***DAN: I think I do not need this section anymore, but just keeping it around to be safe.
% \section{      Fig. \ref{MT-fig:fig4} does not depend on the value of $q_{ij}$ used}\label{sect:F4andqij}
% 
% In Fig. \ref{MT-fig:fig4} of the main text we plotted GWR , $\Delta_i^{[EC]}$, and $\text{GWR}-\Delta_i^{[EC]}$ 
% (see equation \ref{eq:decomp_overall_delta} and the text below it) for the lottery model. We here show
% the resulting plot does not depend on which of the two choices for the scaling factors $q_{ij}$ is used.
% But $\text{GWR} = \olr_{i \bs i} - q_{ij} \olr_{j \bs i}$ and $\Delta_i^{[EC]}=\olep_i^{[EC]}-q_{ij}\olep_j^{[EC]}$,
% so it suffices to show $\olr_{j \bs i}=0$ and $\olep_j^{[EC]}=0$. The first of these is known true, \emph{a priori}, 
% and the second was demonstrated in section \ref{sect:math_simplifications}.

@@

%***DAN: Jasmin, please fix the collision of text that occurs in the text near the legend
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLLNqij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig3} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig3} of the main text, but using 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3qij,5)}, so are not displayed.}\label{fig:fig3_alt}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLBqij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig4} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig4} of the main text, but using 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4qij,5)}, so are not displayed.}\label{fig:fig4_alt}
%***DAN: jasmin, once the se is generated and saved by your code, the chunk at the outset of this 
%document needs to be modified to load that quantity, and then the 0 in the above expression 
%needs to be replaced by the variable name used.
%***JAS:done
\end{figure}

>>>>

\section{Removing asymmetries of association}\label{sect:remove_asymmetries}

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? Fig. \ref{fig:pedagog2} provides an illustration of the steps we describe here. 
We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal (Fig. \ref{fig:pedagog2}a,b).  
Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\E(e_i)=\E(c_{i \bs i})=0$ and $\var(e_i)=\var(c_{i \bs i})=1$, and we denote 
$\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$ (Fig. \ref{fig:pedagog2}c). We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$ (Fig. \ref{fig:pedagog2}d). 
The value of $\ddot{\rho}_i$ is then determined in one of 
four alternative ways: either by setting $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated.
For those familiar with copula statistics (see, e.g., \citealp{Genest_2007} and \citealp{Ghosh_2020_AER}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created similarly.

@@

@@

@@

@@

%***DAN: This looks pretty good except one of the histograms on c does not look like the plotted
%pdfs of the marginals. I think Jasmin said in an email she was fixing this.
%***DAN: Jasmin, also, we use \varphi in the text, but \phi here. These need to be made consistent, since they are different to
%a math person. The easiest way will be to change it in the figure, so I hope that can straightforwardly be done.
\begin{figure}
\includegraphics[width=1.025\textwidth]{../results_figs/theory_fig_col.pdf}
\caption[Visual guide to removing asymmetries of association]{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATA present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{||}$ and  $C_{i \bs i}^{||}$,
%Dan replaced \nat in the above two expressions by || because for some reason the \nat does not
%compile when it occurs within a caption. Not sure how to deal with that, but leaving it alone for now.
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
\label{fig:pedagog2}}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLLNqij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig3} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig3} of the main text, but using 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3qij,5)}, so are not displayed.}\label{fig:fig3_alt}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLBqij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig4} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig4} of the main text, but using 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4qij,5)}, so are not displayed.}\label{fig:fig4_alt}
\end{figure}
%***DAN: Jasmin, the value of se4 quoted in the main text was the same, to 5 digits, as the 
%value of se4qij quoted here. This made me suspicious so I checked the values to 9 digits
%and they were the same to 9 digits. This seems like it must be an error in the code. Please
%check. The standard error expressions in SI S5.2 DO depend on qij, so this seems like it
%should not happen. The standard errors listed in Fig. 3 and Fig. S1 are different, so that's
%good.
